---
title: "Exploratory Analysis"
author: "Frances Hung"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
subtitle: Tables, Plots, and Summary Statistics in R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, fig.width = 8, fig.height = 5)
library(tidyverse)
library(Gmisc)
library(Hmisc)
library(gtsummary)
library(table1)
library(medicaldata)
# library(nhanesA)
#library(flextable)
```

Welcome to the preliminary coding workshops on Quantitative Methods for HIV Researchers (Rishi will send title)! These lectures will serve as code introductions and guides for the in-person portion of the workshop since data analysis and programming will not be covered in the in-person workshops. I will start each of these lectures with a list of packages used and concepts these packages are relevant for.

Today, we will go over creating tables, plots, and summary statistics from tidy data.

# Packages Used Today

- base R (loaded by default and used for any coding in R)
    + summary
    + mean
    + median
    + quantile
- `tidyverse` for data exploration (specifically, `dplyr` for wrangling and `ggplot2` for data visualization)
    + select: select column by name
    + rename\_with: 
    + ggplot
    + geom\_bar
    + labs
- `gtsummary` for tables

Now that we're acquainted with Rstudio and the basic syntax of R, we can briefly talk about exploring data using summary statistics, tables, and visualizations. We will use `dplyr` and `ggplot2`, both which are packages included in the tidyverse bundle of packages.

## Review

Let's clean up the `strep_tb` dataset so we can make tables and figures.

First, we take a quick look at it using the `summary` command.

```{r lwhiv, options}
summary(strep_tb)
```

The values for a lot of the variables consist of a number, underscore, description (e.g. 1\_Good). To avoid redundancy, we remove the number and underscore from those values.

```{r strep-clean, options}
strep_tb_clean <- strep_tb %>%
  # across the five chosen columns
  mutate(across(c(baseline_condition, baseline_temp, 
                baseline_esr, strep_resistance,
                radiologic_6m),
        # remove "_" and whatever comes before it
         ~str_remove(.x, "(.*)_") %>%
        # and make the variable into a factor
           as.factor()))
```

## Data Summary

We first go over how to make descriptive tables and figures, then pivot to summary statistics. We'll be using the `strep_tb` dataset. 

## Tables

If your dataframe is in a tidy format, outputting a Table 1 is fairly straightforward. We use the `tbl_summary` function from the `gtsummary` package to help us create these sorts of tables. The `as_flex_table` function converts the table to a Word-compatible format.

We pipe in the dataframe we want to summarize into the `tbl_summary` function, which also takes in the column names we want to display as well as the column we want to stratify by. If we don't have any stratifying variables in our Table 1, we create a column with one value and "stratify" by that column. 

```{r tables, options}
strep_tb_clean %>%
  # don't include patient ID in table
  select(-patient_id) %>%
  tbl_summary() %>%
  # allows us to knit tables to Word
  as_flex_table()
```

In order to make the variable names look nicer, we can relabel the base dataframe:

```{r re-label, options}
strep_tb_clean %>%
  # don't include patient ID and dose_PAS_g in table
  select(-c(patient_id, dose_PAS_g)) %>%
  # rename all variables by replacing underscores with spaces
  rename_with(function(x) str_replace_all(x,"_", " ") %>%
                # and making everything title case
                str_to_title()) %>%
  tbl_summary() %>%
  # allows us to knit tables to Word
  as_flex_table()
```

We can give the function more optional parameters to make the table 1 more readable (see documentation). Below, we order the factor levels in each factor variable to make them appear in the correct order.

```{r add-params, options}
strep_tb_clean %>%
  # don't include patient ID and dose_PAS_g in table
  select(-c(patient_id, dose_PAS_g)) %>%
  # re-level strep_resistance, baseline_temp, baseline_condition, radiologic_6m
  mutate(strep_resistance = factor(strep_resistance,
                                   levels = c("0-8", "8-99", "100+")),
         baseline_temp = factor(baseline_temp,
                                levels = c("98-98.9F", "99-99.9F",
                                           "100-100.9F", "100F+")),
         baseline_condition = factor(baseline_condition,
                                     levels = c("Poor", "Fair", "Good")),
         radiologic_6m = factor(radiologic_6m %>% str_to_title(),
                                levels = c("Change", "Improvement", "Deterioration", "Death"))) %>%
  # rename all variables by replacing underscores with spaces
  rename_with(function(x) str_replace_all(x,"_", " ") %>%
                # and making everything title case
                str_to_title()) %>%
  tbl_summary() %>%
  # allows us to knit tables to Word
  as_flex_table()
```

## Visualizations

In order to visualize our data, we can use the `ggplot2` package, which is loaded in through tidyverse. Plotting requires feeding a dataframe with the variables we want to plot into the `ggplot` function, along with the column names of the variables we want plotted on the x and/or y axis. 

In the example below, the parameters inside of aes() tells `ggplot2` tells it information essential for mapping data to the plot. The x axis should consist of the treatment arm, and whatever plot we make should be colored (filled) by the radiographic number.

```{r empty-ggplot, options}
strep_tb_clean %>%
  ggplot(aes(x=arm, fill=as.factor(rad_num)))
```

Notice that nothing was plotted when the above code chunk was run. This is because we then have to tell ggplot what sort of graph (geoms) or graph component we'd like to add. For example, we can plot a barplot which has arm on the x-axis and viral load on the y-axis by adding `geom_boxplot()` to the base ggplot layer. We add the "fill" aesthetic to specify that the barplot should be filled (colored) in by radiographic number. 

```{r label, options}
strep_tb_clean %>%
  ggplot(aes(x=arm, fill=as.factor(rad_num))) +
  geom_bar(position = "dodge") 
```

The `ggplot2` syntax is additive, so we can keep on adding graphs and graph components. Below, we add points and axis labels to replace the current ones:

```{r more-ggplot-layers, options}
strep_tb_clean %>%
  ggplot(aes(x=arm, fill=as.factor(rad_num))) +
  geom_bar(position = "dodge") +
  labs(fill="Radiographic Outcome \n (Worst-Best)",
       x="Treatment Arm",
       y="# Patients")
```

### More Geoms

There are many more geoms that `ggplot2` supports. A few commonly used ones include:

- Histograms: geom_histogram()
- Boxplots: geom_boxplot()
- Points: geom_point()

Remember that some geoms need more information than others. For boxplots for example, you need to provide both x and y parameters in the aes() argument. 

The below boxplot has the same information as the barplot above, but we treat radiographic number as a continuous variable.

```{r geom-boxplot, options}
strep_tb_clean %>%
  ggplot(aes(x=arm, y=rad_num)) +
  geom_boxplot() +
  labs(x="Treatment Arm",
       y="Radiographic Outcome")
```

### Exercises: Plots

1. Add data points to the boxplot graph. What improvements could be made to the graph?
2. Look up `geom_jitter` and use it instead to add points to the boxplot graph.
3. Look up `geom_bar` and use it to plot a bar graph of strep resistance, colored by whether the patient improved.

```{r geom-bar, options}
strep_tb_clean %>%
  mutate(strep_resistance=factor(strep_resistance,
                                 levels = c("0-8", "8-99", "100+"))) %>%
  ggplot(aes(x=strep_resistance, fill=improved)) +
  geom_bar() +
  labs(x="Strep Resistance", y="Count", fill="Improved")
```

## Summary Statistics

### Foundational summary statistics

```{r foundation-stats, options}
# the mean
mean(strep_tb_clean$rad_num)

# the median
median(strep_tb_clean$rad_num)

# the 20% and 80% quantiles
quantile(strep_tb_clean$rad_num, c(0.2, 0.8))
```

### Summary Statistics by Group

We went over functions for mean, median, and quantiles of dataframe columns, but we often want these summary statistics by group. For example, patients in the TB study had different starting conditions; their radiographic outcomes may vary by starting condition.

Two paired functions are useful for summary statistics: `group_by` and `summarise`. The first tells us what groups we want to summarize the data by, and the second tells us what function we want to use to summarize.

```{r condition_summarise, options}
strep_tb_clean %>%
  # group by baseline condition
  group_by(baseline_condition) %>%
  # find mean by the above baseline condition groups
  dplyr::summarize(meanRadNum=mean(rad_num))
```

A short cut for a popular summary statistic is `count`. The arguments, like for `group_by`, designate the groups we want to summarize the data by.


```{r condition-count, options}
strep_tb_clean %>%
  group_by(baseline_condition) %>%
  dplyr::summarise(n=n())

strep_tb_clean %>%
  count(baseline_condition)
```


## Exercises

1. Summarize how many male and female (`gender`) patients are in each treatment arm (`arm`).
2. Find the median radiological number (`rad_num`) by gender and treatment arm.

```{r summary-stat-ex, options}
strep_tb_clean %>% count(arm, gender)
strep_tb_clean %>% group_by(arm, gender) %>% summarise(medRad=median(rad_num))
```
